---
title: 'Week 7: Dates and Strings'
author: "Ellen Bledsoe"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Working with Dates and Times

This week, we are going to learn a bit about how to work with some of the more troublesome data types in R: dates and character strings.

## Set-Up

Fortunately, the `tidyverse` includes packages to help us with both.

```{r}
library(tidyverse)
```

We are going to use some stream data from the Santa Cruz River here in Tucson. The data are recorded every 15 minutes; these data are from 2023.

```{r}
download.file("https://nwis.waterservices.usgs.gov/nwis/iv/?sites=09482495&parameterCd=00060&startDT=2023-01-01T00:00:00.000-07:00&endDT=2023-12-31T23:59:59.999-07:00&siteStatus=all&format=rdb", "santa_cruz.txt")
```

When we take a look at the data, we can see a number of rows that we need to skip when we read in the data...27 to be exact.

```{r}
water <- read_tsv("santa_cruz.txt", skip = 27, col_names = TRUE)

head(water)
```

Ok, we are close but not quite there. The first row of data is metadata rather than data.

To remove that, we are going to use a new function called `slice()`. This removes rows from a dataframe based on position. We want to remove the first row.

```{r}
water <- water %>% 
  slice(-1)

head(water)
```

That's better! Let's also rename the last 2 columns. Those numbers are going to be pretty annoying to deal with down the road.

The stream discharge column should be numeric but got read in as character for some reason, so we need to convert it to a numeric data class.

```{r}
water <- water %>% 
  rename(discharge_ft3_s = `307121_00060`,
         qualification_code = `307121_00060_cd`) %>% 
  mutate(discharge_ft3_s = as.double(discharge_ft3_s))
water
```

Let's try to plot the amount of water flowing through the stream through time.

```{r}
ggplot(water, aes(datetime, discharge_ft3_s)) +
  geom_point()
```

Well, that doesn't look great...

As it turns out, the `datetime` column is also currently character data. In order to use this column efficiently, we will want to convert it to a date (or datetime) data class.

## Dates and Times with `lubridate`

Dates and times can be particularly challenging to work with, but thankfully the `lubridate` package makes it a bit easier.

### Referencing the Current Date and Time

First, let's start with functions that return the current date and/or time. These are helpful if you want to "stamp" with the date/time something happens (e.g., when exactly did you knit that PDF file).

```{r}
# date
today()

# date and time
now()
```

Aside from these functions, we will mostly be using the `lubridate` functions within a `mutate` function to modify entire columns at once.

### `datetime`

The most generic function in `lubridate` is `as_date` or the variant, `as_datetime`. These convert non-dates into a date data class.

We need to specify the following arguments:

-   the datetime data to be converted to a date
-   the format of the datetime data
-   optionally, we can specify the time zone

The way we specify the format of the datetime is a bit unusual. We use "conversion specifications," which are introduced by a `%` and are typically represented by a letter. There are lots of options, but the ones we will use are below:

-   `%Y`: 4 digit year
-   `%y`: 2 digit year
-   `%m`: month in numeric form
-   `%d`: day of month in numeric form
-   `%H`: hours in 24-hour format
-   `%M`: minutes
-   `%S`: seconds

The characters that are used to separate the date and time data are interpreted literally. What does that mean? If the date is separated by `-`, the conversion specifications should be separated by `-`, too.

```{r}
water %>% 
  mutate(datetime = as_datetime(datetime, format = "%Y-%m-%d %H:%M", tz = "MST"))
```

We can now see that the `datetime` column is something new, called `POSIXct`. 

You don't need to know too much about what that means. Behind the scenes, the date/time is being stored as the number of seconds since Jan 1, 1970 at 00:00:00 UTC, allowing R to perform mathematical calculations with dates and times.

```{r}
max(water$datetime)
```

### Functions Specifying Date/Time Structure

Any time I try to remember the conversion specifications, I draw a blank. Are they uppercase or lowercase? Some letters don't match the first letter of the data it represents.

`lubridate` provides functions with nearly every iteration and combination of year, month, day, hour, minutes, and seconds. 

When we use these, we don't need to specify the exact format or the separators, which is particularly helpful if you have dates with different separators in the same column.

As we know from the previous code, our `datetime` column has year, month, day, hours and minutes.

```{r}
water <- water %>% 
  mutate(datetime = ymd_hm(datetime))
water
```

Now, when we plot the trend of water discharge through 2023, we can see that ggplot knows how to deal with the values more efficiently.

```{r}
ggplot(water, aes(datetime, discharge_ft3_s)) +
  geom_point()

```
### Extracting Specific Components

Sometimes, we don't want all of the data from the datetime or we want to `filter`, `group_by`, `summarize`, etc. by a part of the datetime.

For example, perhaps we want to know the average water discharge per month. We would first want to create a new column with the month for each observation and then `group_by` and `summarize`.

There are numerous functions we can use to extract specific components from the datetime.

We can use the `date()` function to pull out the date from the datetime. As a note, it is surprisingly challenging to create a time column...

```{r}
water %>% 
  mutate(Date = date(datetime))
```

Let's create columns for each segment.
```{r}
water <- water %>% 
  mutate(Year = year(datetime),
         Month = month(datetime, label = TRUE),
         Day = day(datetime),
         Hour = hour(datetime),
         Minutes = minute(datetime),
         Seconds = second(datetime))
water
```

Now, we can calculate the average value per month.
```{r}
water %>% 
  group_by(Month) %>% 
  summarise(mean = mean(discharge_ft3_s))
```

We can also create new columns such as day of year (DOY). This is the day of year, running from 0-365. One common use for DOY is in phenology, measuring the annual timing of life history stages. The function to calculate DOY is `yday()`.

```{r}
water %>% 
  mutate(DOY = yday(datetime), .after = tz_cd)
```

### Creating Dates from Segments

What if you are given a dataset that has separate year, month, and day columns that you need to combine? If you recall, this is how the data frame for the Portal rodents is set up.

```{r}
surveys <- read_csv("../Week4_Groups_and_Joins/surveys.csv")
surveys
```

We can bring these columns together to form a date column using the `make_date()` function.

```{r}
surveys <- surveys %>% 
  mutate(date = make_date(year = year,
                          month = month,
                          day = day))
surveys
```

### Performing Calculations

Because of the unique way that dates and datetimes are stored in R, we can also perform calculations with them.

In particular, `lubridate` has functions to calculate intervals between dates and datetimes.

```{r}
start <- min(surveys$date, na.rm = TRUE)
end <- max(surveys$date, na.rm = TRUE)

date_interval <- interval(start, end)
date_interval

as.duration(date_interval)
```

