---
title: "TEMP: Challenge Question for the Dates and Strings Assignment"
format: pdf
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# the assignment loads the tidyverse in Question 1; it is loaded here so this file runs on its own
library(tidyverse)
```

This is a temporary file. Everything under the first heading is the question, ready to paste into the assignment as an additional (optional) question. The instructor solution is at the bottom; delete it before using the question.

# Question to Add

### 8. Challenge: What Do Kangaroo Rats Eat? (Optional)

*This challenge question was mostly created by Claude, an AI model from Anthropic, with direction and revisions from Ellen Bledsoe.*

During my PhD, I collected fecal samples from rodents at the Portal Project and used DNA metabarcoding to figure out which plants they had eaten. Let's wrangle some of those data!

Run the code chunk below to read in two data frames.

- `fecal` has one row per fecal sample, including the month, day, and year it was collected and a `vial_barcode` that identifies the sample.
- `reads` has the DNA sequencing results for the kangaroo rat samples. Each row is one type of DNA sequence (`OTU`) found in one sample (`SampleID`, which matches `vial_barcode`). `Reads` is the number of times that sequence was detected. `ConsensusLineage` is the taxonomy assigned to the sequence, stored as one long string in which `f__`, `g__`, and `s__` come before the family, genus, and species (e.g., `k__Eukaryota; p__Streptophyta; c__Liliopsida; o__Poales; f__Poaceae; g__Panicum; s__Panicum miliaceum`).

```{r}
fecal <- read_csv("https://raw.githubusercontent.com/weecology/DNA_metabarcoding/master/data/collection_data/fecal_sample_collection.csv")

reads <- read_csv("https://raw.githubusercontent.com/weecology/DNA_metabarcoding/master/data/sequenced_data/krat_fecal_its.csv") %>%
  select(SampleID, OTU, Reads, ConsensusLineage)
```

a.  Using `make_date()` and `mutate()`, add a `collection_date` column to `fecal`. Then use `month()` (with `label = TRUE`) to add a `month_name` column. Save the result as `fecal`.
b.  The `reads` data frame contains more than plants: some sequences are bacteria or lab contaminants. Use `filter()` and `str_detect()` to keep only the sequences whose lineage contains "Streptophyta" (the green plants). In the same `filter()`, also remove the sequences from millet (*Panicum miliaceum*), which was the bait in the traps and not a natural food. Save the result as `plant_reads`. *(Hint: `!` reverses a TRUE/FALSE vector.)*
c.  Use `mutate()` and `str_extract()` to create a `family` column in `plant_reads` containing the plant family from `ConsensusLineage` (the text after `f__` and before the next `;`). Use the regex pattern `"(?<=f__)[^;]+"`. Some sequences were not identified to a family and have the value "None"; convert those to `NA` and then drop those rows. Save the result as `plant_reads`.
d.  Join `fecal` and `plant_reads`, keeping only the samples that are in both (the sample ID column has a different name in each data frame). Then calculate the total number of reads for each plant family in each month, and the proportion of each month's reads that came from each family. Save the result as `family_by_month`.
e.  Reshape `family_by_month` so there is one row per family and one column per month, with the proportions in the cells. Fill any family-month combination with no reads with 0, and sort the rows so the family with the highest proportion in March is at the top.
f.  Make a plot of the proportion of reads for each plant family, with the family on the x-axis, the proportion of reads on the y-axis, and the months shown in different colors. To keep the plot readable, include only the families that make up at least 10% of the reads in at least one month. Make the points size 3, label the axes clearly, and add a theme. *(Hint: Adding `theme(axis.text.x = element_text(angle = 45, hjust = 1))` makes the family names easier to read.)*

```{r}

```

# Instructor Solution (delete before using)

```{r}
fecal <- read_csv("https://raw.githubusercontent.com/weecology/DNA_metabarcoding/master/data/collection_data/fecal_sample_collection.csv")

reads <- read_csv("https://raw.githubusercontent.com/weecology/DNA_metabarcoding/master/data/sequenced_data/krat_fecal_its.csv") %>%
  select(SampleID, OTU, Reads, ConsensusLineage)
```

8a. The collection dates fall in four months; only three of them (Oct, Mar, Sep) have sequenced samples.

```{r}
fecal <- fecal %>%
  mutate(collection_date = make_date(year, month, day),
         month_name = month(collection_date, label = TRUE))

fecal %>%
  distinct(collection_date, month_name)
```

8b. `str_detect()` returns TRUE/FALSE, so it works directly inside `filter()`; `!` flips it to remove the millet.

```{r}
plant_reads <- reads %>%
  filter(str_detect(ConsensusLineage, "Streptophyta"),
         !str_detect(ConsensusLineage, "Panicum miliaceum"))

plant_reads
```

8c. The lookbehind `(?<=f__)` means "preceded by `f__`" and `[^;]+` means "one or more characters that are not a semicolon." The 3 rows with no family come out as the text "None", which `na_if()` turns into `NA`.

```{r}
plant_reads <- plant_reads %>%
  mutate(family = str_extract(ConsensusLineage, "(?<=f__)[^;]+"),
         family = na_if(family, "None")) %>%
  drop_na(family)

plant_reads
```

8d. An `inner_join()` keeps only the samples that were sequenced. After `summarize()`, the data are still grouped by month, so `mutate()` calculates the proportions within each month.

```{r}
family_by_month <- plant_reads %>%
  inner_join(fecal, join_by(SampleID == vial_barcode)) %>%
  group_by(month_name, family) %>%
  summarize(total_reads = sum(Reads)) %>%
  mutate(prop_reads = total_reads / sum(total_reads)) %>%
  ungroup()

family_by_month
```

8e.

```{r}
family_wide <- family_by_month %>%
  select(month_name, family, prop_reads) %>%
  pivot_wider(names_from = month_name,
              values_from = prop_reads,
              values_fill = 0) %>%
  arrange(desc(Mar))

family_wide
```

8f. Seven families make up at least 10% of the reads in at least one month. Filtering the wide table is simple (`|` means "or"), and `pivot_longer()` then puts the data back into long format for `ggplot()`. (Alternatively, `family_by_month %>% group_by(family) %>% filter(max(prop_reads) >= 0.1) %>% ungroup()` gives the same families.)

```{r}
common_families <- family_wide %>%
  filter(Mar >= 0.1 | Sep >= 0.1 | Oct >= 0.1) %>%
  pivot_longer(Mar:Oct,
               names_to = "month_name",
               values_to = "prop_reads")

ggplot(common_families, aes(x = family, y = prop_reads, color = month_name)) +
  geom_point(size = 3) +
  labs(x = "Plant family",
       y = "Proportion of reads",
       color = "Month") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
