---
title: "Assignment 7, Part I"
author: "Ellen Bledsoe"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Assignment Details

### Purpose

The goal of this assignment is to work with dates and times using the `lubridate` package.

### Task

Write R code to successfully answer each question below.

### Criteria for Success

-   Code is within the provided code chunks or new code chunks are created where necessary
-   Code chunks run without errors
-   Code chunks have brief comments indicating which code is answering which part of the question
-   Code will be assessed as follows:
    -   Produces the correct answer using the requested approach: 100%
    -   Generally uses the right approach, but a minor mistake results in an incorrect answer: 90%
    -   Attempts to solve the problem and makes some progress using the core concept, but returns the wrong answer and does not demonstrate comfort with the core concept: 50%
    -   Answer demonstrates a lack of understanding of the core concept: 0%
-   Any questions requiring written answers are answered with sufficient detail

### Due Date

March 11 at midnight MST

# Assignment Exercises

The assignment for week 7 is divided into 2 parts:

-   Part 1: `lubridate`
-   Part 2: `stringr`

### 1. Set-Up (5 pts)

Load in the `tidyverse`.

### 2. When Did You Knit This Document? (5 pts)

Write code to do the following:

a.  The first line should print only the date that you knit this document.
b.  The second line should print the date and time that you knit this document.

### 3. Plant Vouchers (20 pts)

During my PhD, I collected plant vouchers from my field site that I eventually submitted to the UA Herbarium for identification. I also got DNA sequences for most of them (but not all). Read in that dataset by running the code chunk below.

```{r}
vouchers <- read_csv("https://raw.githubusercontent.com/weecology/DNA_metabarcoding/master/data/collection_data/plant_voucher_collection.csv")
vouchers
```

a.  Using the `make_date` function and the `mutate()` function, create a new column called "collection_date" that has the year, month, and day that I collected the voucher specimen for each plant. Save this new column in the `vouchers` dataframe.
b.  Using the `min()` function, find the earliest date that I collected a voucher specimen. Using the `summarize()` function is optional.
c.  Use code to find the last date that I collected a vocuher specimen.
d.  Find the span of time (duration) that I was collecting voucher specimens.
e.  Create a column for the day of year that each specimen was collected.

### 4. NDVI from the Santa Rita Experimental Range (20 pts)

There is a large network of phenocams (cameras set up to take daily images of a landscape to monitor plant phenology) across the US. The Santa Rita Experimental Range (SRER) has one such camera.

One value that we can calculate from these phenocam images is the NDVI of the landscape in the picture, which gives us a measurement of vegetation "greenness." This value is likely to change through time due to seasonal changes in temperature, precipitation, etc.

Run the following code chunk to bring in a subset of the SRER phenocam data.

```{r}
phenocam <- read_csv("neon_srer_ndvi_phencom.csv", skip = 17, col_names = TRUE, 
                    col_types = "ccncc") %>% 
  select(date, local_std_time, contains("mean"), NDVI_c)
```

a.  Create a new column in `phenocam` that uses the `unite()` function to join the date and time columns together into one column. Separate the date and time with a space.
b.  Convert the datetime column you created in (a) to a POSIXct format. Save this new column to `phenocam`. You can use whichever function you would like.
c.  Calculate the duration of the phenocam dataset provided here.
d.  Create new columns in `phenocam` for the year, the month, and the day of year.
e.  Using the year and month columns you created in (d), calculate the average NDVI value for each month in the dataset (each month in each year).
