---
title: 'Week 9: Version Control'
author: "Ellen Bledsoe"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Reproducibility

## Version Control

### git and GitHub

### Using git and GitHub in RStudio

## Introduction

### Motivation

Get rid of messy folders and track changes to things like data files and code in a more manageable way.

### Benefits of version control

-   Track changes (*but better*)
    -   Tracks every change ever made in groups called commits
        -   Every commit stores the full state of all of your files at that time
    -   Never lose anything
        -   Revert or restore to any commit
        -   Easily unbreak your code/data/manuscript
        -   No more file name changes
-   Collaboration
    -   Work on things simultaneously
        -   See what changes others have made
        -   Everyone has the most recent version of everything

## Version control using Git & RStudio

### Create a Git repo

1.  Navigate to Github in a web browser and login.
2.  Click the `+` at the upper right corner of the page and choose `New repository`.
3.  Choose the class organization (e.g., `dcsemester`) as the `Owner` of the repo.
4.  Fill in a `Repository name` that follows the form `FirstnameLastname`.
5.  Select `Private`.
6.  Select `Initialize this repository with a README`.
7.  Click `Create Repository`.

### Connect to the Git repo in RStudio

1.  From new GitHub repository, click green `Clone or download` button -\> Click the `Copy to clipboard` button.
2.  In RStudio, File -\> New Project -\> Version Control -\> Git
3.  Paste copied URL in `Repository URL:`.
4.  Leave `Project directory name:` blank; automatically given repo name.
5.  Choose where to `Create project as subdirectory of:`.
6.  Click `Create Project`.
7.  Check to make sure you have a `Git` tab in the upper right window.

### Install

```{r}
#install.packages("usethis")
library(usethis)
```

### Introduce yourself to Git

```{r}
library(usethis)

use_git_config(user.name = "[name]", user.email = "[email]")
```

### First commits

#### Commit data

-   Download the data file [Gaeta_etal_CLC_data.csv](%7B%7B%20site.baseurl%20%7D%7D/data/Gaeta_etal_CLC_data.csv) to your project directory.
-   Add the data file to version control
-   Two step process:

1.  Add the data file (checkbox)
2.  Commit it

-   Git -\> Select `Gaeta_etal_CLC_data.csv`.
-   Commit with message.
    -   `Add fish size and growth rate data`
-   History:
    -   One commit
    -   Changes too large to see

#### Commit R script

-   Read in data to new R script.

```{r}
library(tidyverse)

fish_data <- read_csv("Gaeta_etal_CLC_data.csv")
```

-   Save as `fish-analysis.qmd`.
-   Git -\> Select `fish-analysis.qmd`.
    -   Changes in staged files will be included in next commit.
    -   Can also see changes by selecting `Diff`
-   Commit with message.
    -   `Start script comparing fish length and scale size`
-   History:
    -   Two commits
    -   See what changes were made to `fish-analysis.qmd`

### Building a history

-   `fish-analysis.qmd` doesn't currently show on the `Git` tab
    -   No saved changes since last commit
-   Add some more code to `fish-analysis.R`
    -   Create new categorical size column

```{r}
fish_data_cat = fish_data |>
  mutate(length_cat = if_else(length > 200, "big", "small"))
```

-   Save `fish-analysis.qmd`.
-   Now we see the file on the `Git` tab.
    -   `M` indicates that it's been modified.
-   To commit these changes, we need to stage the file.
    -   Check the box next to `fish-analysis.R`.
-   Commit with message.
    -   `Add categorical fish length column`
-   History:
    -   Three commits
    -   Each `fish-analysis.R` commit shows the additions we made in that commit.
-   Modify this code in `fish-analysis.R`
    -   Change category cut-off size

```{r}
fish_data_cat = fish_data %>% 
  mutate(length_cat = if_else(length > 300, "big", "small"))
```

-   Save file -\> stage -\> commit
    -   `Change size cutoff for new column`
    -   Green sections for added lines, red for deleted
    -   Git works line by line.
        -   The previous version of the line is shown as deleted.
        -   The new version of the line is shown as added.

### Committing multiple files

-   Commits can include multiple files at once
-   Let's move our data file into a `data` subdirectory
-   `New Folder` -\> `data`
-   Checkbox `Gaeta_etal_CLC_data.csv` -\> `More` -\> `Move`
-   Change code to read from new subdirectory

```{r}
fish_data = read.csv("data/Gaeta_etal_CLC_data.csv")
```

-   Changes to R script indicated by M

-   Original datafile has a red D next to it which indicates "deleted"

-   New, untracked, data directory

-   git initially thinks we've deleted `Gaeta_etal_CLC_data.csv` and created a new `Gaeta_etal_CLC_data.csv` file in a new directory.

-   Click on both the old and new files to stage them

-   git then recognizes that we have moved (or renamed) the file by making the two files into one and marking this with an `R` for "rename".

-   Commit: `Move data file into subdirectory`

### Git as a time machine

#### Experiment with impunity

```{r}
fish_data_cat = fish_data %>% 
  mutate(length_cat = ifelse(length > 300, <b>"large"</b>, "small"))
```

-   `Save` and show changes are staged

-   `More` -\> `Revert` -\> `Yes`

-   Get previous state of a file

    -   `History` -\> select commit -\> `View file @ ...`
    -   Save file over current file
    -   Copy and paste relevant piece into current file

#### Delete with impunity

-   Both of these also work for deleted files
-   Close the upper left window with the `fish-analysis.R`.
-   Choose the `File` tab in the lower right window.
-   Select `fish-analysis.R` -\> `Delete` -\> `Yes`
-   Stage deleted file -\> <i class="fa fa-gear"></i> `More` -\> `Revert` -\> `Yes`

## GitHub Remotes

-   So far we've worked with a local `Git` repository.
-   One of the big benefits of version control is easy collaboration.
-   To do this, we synchronize our local changes with a remote repository called `origin`.
-   Our remote repository is on GitHub.
    -   By far the most popular hosted version control site
    -   Public and private hosted repositories
    -   Private free for students and academics
    -   <https://education.github.com/>
        -   For the assignment, we're using private repositories that we made at the beginning.

### Push to a remote

#### Connect to GitHub

-   To push to your remote we first have to connect to GitHub, which is a little tricky
-   First, log in to GitHub in your browser
-   Then create a GitHub token, this is like a special password just for one computer

```{r}
usethis::create_github_token()
```

-   Select defaults

-   Create token

-   Copy token

-   Now add this token our local git setup so that it can use it to connect to GitHub

```{r}
gitcreds::gitcreds_set()
```

-   Paste your password

#### Push

-   `Push` sends your recent commits to the `origin` remote.
-   Before a `Push` your commits show in your local history but not on the remote.
-   To `Push` to your remote, select the `Push` button at the top of the `Git` tab.
-   Now your changes and commit history are also stored on the remote.

```{r}
ggplot(fish_data_cat, aes(x = scalelength, fill = length_cat)) +
  geom_histogram()
```

> Either you (logged in as another user) or your teaching partner should make the same change to your respository

### Pulling

-   Big advantage to remotes is easy collaboration

-   Avoids emailing files and shared folders where you are never sure if you actually have the most recent version

-   Makes it easy to see what collaborators have done

-   Automatically combines non-overlapping changes

-   While I've been talking, a collaborator has added a plot of scale size and fish length to the code.

-   `Pull` the changes from the remote repo with the `Pull` button on the Git tab

### Merges

> Demo merges either with a partner or by logging into GitHub as another user in the browser.

-   What happens if two people make changes at the same time?

    -   If they edit different parts of the code git will combine them automatically
    -   If they edit the same areas of the code this requires human intervention

-   Merges

-   You decide to change the number of histogram bins to 10

```{r}
geom_histogram(bins = 10)
```

-   Your collaborator reassesses the measurement device and decides it is accurate down to 0.5 mm and pushes the change to the remote repository [make this change in the remote]

```{r}
filter(scalelength >= 0.5)
```

-   You try to push your change
-   Get an error that shows someone else has made a change & you need to incorporate it to push
-   Pull
-   Merge happens automatically
-   You have both sets of changes
-   Remote still only has collaborators changes
-   Push to add the merged version to the remote

### Merge conflicts

-   If both you and your collaborator edit the same location in the code git doesn't know how to combine the changes.

-   A human has to make this kind of decision.

-   You decide to change `"big"` to `"large"`

```{r}
mutate(length_cat = ifelse(length > 300, <b>"large"</b>, "small"))
```

-   Your collaborator changes the size threshold and pushes to the remote

```{r}
mutate(length_cat = ifelse(length > 250, <b>"big"</b>, "small"))
```

-   You attempt to push your changes
-   Merge conflict when pulling collaborators changes
-   This shows as `U` for "unmerged" in RStudio
-   First block of code is your version
-   Second block is the version on the remote
-   Combine into a single block that includes everything

```{r}
mutate(length_cat = ifelse(length > 250, "large", "small"))
```

-   Click check box next to file
-   Commit indicating that it is a merge
-   Still not on remote yet
-   Push
