---
title: 'Week 9: Reproducibility in RStudio'
author: "Ellen Bledsoe"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Reproducibility in RStudio

In this week's lesson, we will be talking broadly about how to make our data analyses reproducible.

We will do this through 2 different approaches. 1) Reproducibility in RStudio 2) Using git/GitHub for version control

For today's lesson, we will focus on how we can increase reproducibility within RStudio.

## R vs. RStudio

Before we begin, it is worth reminding ourselves of the difference between R and RStudio.

We are making the shift from working in Posit Cloud to using R and RStudio on your own device, meaning you should have both R and RStudio downloaded on your computer.

For most of us, even though we have downloaded the R software, we will likely never interact with the R program directly; I personally never open up R on my own computer.

Instead, in order to code in R and work with the R software, we will want to open up RStudio. Like it does in Posit Cloud, RStudio on our own computers connects to the R software that we have downloaded and let's us interact with R within the RStudio environment. So while I never personally open up R software directly, I frequently open up RStudio and work with R there!

## Introduction to Reproducibility

Throughout the course, we have talked about many of our topics in the context of "reproducbility," but what *exactly* does that mean?

When something is *reproducible*, it means we have the ability to recreate the same results (including cleaned data, tables, figures, and quantitative findings), using the same input data, computational methods, and conditions of analysis.

Our ultimate goal is to be able to rerun a fully analysis with a single click (or command).

Our first step is to make sure our code runs anytime and anywhere. That includes the next day or 5 years from now, on a desktop or a laptop, a Windows or a Mac, or on a collaborator or advisor's computer.

Below are some guidelines for making this happen:

**1. Make sure things you did before don't matter (if they shouldn't)**

Computers store the results of each command that we run in sequence. Occasionally, we will make a change to the code (perhaps we are testing a complicated piece of code) and the code will look like it still runs.

Sometimes, that line of code only works because of something we did earlier in the session, but if we restarted the session, it would get stuck because an object doesn't get created in the code or is created later on in the code.

We have actually been dealing with this scenario already. When we "knit" a .Rmd file--into a PDF, in our case--the document runs from scratch. If there is something in the code that is out of order or an object that isn't created in the code, the RMarkdown file won't create the PDF.

**2. Clearing environments and restarting R**

We can check that our code runs without having to knit, though. We have a few options.

-   Clear R environment using the broom icon on the `Environment` tab.

    -   Doesn't unload packages
    -   Useful when developing code

-   Restart R to get a clean environment

    -   Does unload packages
    -   Useful for making sure everything works
    -   As long is it doesn't secretly reload things

-   Run entire file by selecting "Run All" in RMarkdown (`Ctrl + Alt + R` or `Cmd + Opt + R`)

For all of these options, we are trying to make sure that the code runs fully and produces the desired result without needed any tweaks.

**3.  Stop saving the current state of the environment**

The typical default in RStudio is to automatically save the state of the environment when you close out of RStudio. Alternatively, it might ask if you want to save or not. 

If you choose to save the workspace, everything in the environment will get reloaded when you start R, even when you restart it as described above.

While that might sound convenient, it can actually cause a lot of problems (see #1 above). To reduce the risk of having code that doesn't fully run on one click, we want to stop R from storing the state of the environment when we close RStudio.

To set the default in RStudio to not save the workspace, we want to do the following:
`Tools` -\> `Global Options` -\> `General` -\> `Save workspace to ~/.RData on exit` -\> `Never`

Unclick `Restore .RData into workspace at startup`

**4.  Make sure code works on other computers** 

We will talk a lot more in-depth about this in a few minutes, but generally speaking, we want to do the following things to increase the likelihood that our code will run on another computer:

-   Avoid using `setwd()` and absolute file paths (e.g., `C:\Users\Batman\DataCarp\data\mydata.csv`)
-   Instead, use R Projects and relative paths (e.g, `data/mydata.csv`)
-   Write code that will work on all operating systems
    -   Filenames in code should match actual names exactly, including capitalization
    -   Use `/` instead of `\` or `\\` in paths

**5.  Clean up extra code**

-   Remove `install.packages()` lines from your code to avoid reinstalling packages repeatedly
-   Remove experiments from your code
-   If you don't want to remove them, comment them out by putting a `#` in front of the line of code.
    - A shortcut for this is `Ctrl + Shift + C` or `Cmd + Shift + C`

## Project Structure

### Introduction

-   Learn how to organize a computational project

-   Many benefits to good organization & documentation

    -   Understand all project components
    -   Work more efficiently
    -   Easier for collaboration
    -   Including with your future self

-   Save a lot of time & frustration in long run, even if it’s time-consuming and frustrating to do it right initially

-   Good file structure

    -   All project files in one main folder
    -   Subfolders (data, R, etc.)

-   Main folder is R project

    -   Self-contained project
    -   Use relative instead of absolute paths

-   Good folder & file names

    -   Descriptive but not too long
    -   No spaces
    -   Consistent format

-   README

    -   Everything someone needs to know to understand project

-   Raw data

    -   In separate folder from cleaned data
    -   Never change!
    -   Each file should have metadata

-   Scripts with code

    -   Relative file paths to read in and create files
    -   Lots of comments
    -   Order: libraries, data, user-created functions, everything else
    -   Good variable & column names

-   Two methods to deal with different versions of same file

    -   Good = consistent naming scheme
        -   Each copy has date or version number
        -   Don’t call anything final, cause no file ever is!
    -   Better = version control, only have one copy of each file

-   Running all scripts

    -   Good = description of order to run each script in README
    -   Better = script to run all scripts
    -   Best = Make

-   Combine code output and text in single document, such as with RMarkdown files (which we are already doing!)

### Goal of good project structure

-   Reproducibility = ability to get same results from set of data & code
-   Give entire project folder to someone else (or your future self!), they can understand all parts and recreate your results, figures, documents, etc.
-   Improving reproducibility makes YOUR life easier!

### R Projects

The best way to work in RStudio is through R Projects (.Rproj files).

### Example project walk-through

Download [example project](%7B%7B%20site.baseurl%20%7D%7D/data/example_project.zip).

### File Paths

Why do I believe so strongly in working in RProjects? There are a few reasons. In part, it is because R Projects work so well with the version control that we will be learning in the next lesson. The other reason is file paths!

### (optional) If Using Posit Cloud

-   So far in this course we've been using Posit Cloud and placing data files where R knows to find them for you
-   Working with data files on your own computer is a little more complicated
-   But in this lesson we'll learn how to do this effectively

### Introduction to Paths

-   To use data stored on a computer we need to tell R where it is

-   This is done using paths

-   This is a description of the directories where our files are stored

-   Let's go to the course website and download some data

-   <https://datacarpentry.org/semester-biology/materials/datasets>

-   This page contains all of the data we use in class

-   Download Shrub Dimensions data

-   If we click on this link it will download, but where did it download to

-   Click on the arrow to 'Show in Folder'

-   The details will look different depending on the operating system

-   But, generally, across the top you'll see the folder that things are stored in

-   Go back to RStudio

-   If we try to load this data just using it's filename it won't work

``` r
data = read.csv("shrub-dimensions-labeled.csv")
```

-   Returns the error "Cannot open file", "No such file or directory"

-   That's because it can't find the file where we've tried to load it from

-   Paths can be either relative or absolute

-   We'll start with absolute, which always describes exactly where something is on the computer

-   Data file is stored in the `Downloads` subdirectory of our `Home` directory

-   `Home` directory varies by operating system

-   On mac and Linux it is `/home/username`

-   Within that is the `Downloads` directory

``` r
# OSX/Linux
data = read.csv('/home/ethan/Downloads/shrub-dimensions-labeled.csv')
```

-   This successfully loads the data because we've told it exactly where the file is
-   On Windows change `home` to `Users`

``` r
# Windows
data = read.csv('/Users/ethan/Downloads/shrub-dimensions-labeled.csv')
```

-   Folders/Directories are separate by `/` with the file name at the end

-   Windows shows `\` or `\\` as the separator, but `/` works on all operating systems so use it

-   Include the file extension (the part after the `.`)

-   If you ever need to go back/up a level, use `..`

-   Paths can also be relative

``` r
'Downloads/shrub-dimensions-labeled.csv'
```

-   "From where I am the shrub-dimensions file is in the Downloads subdirectory"
-   The absolute & relative paths here are the same if R thinks it's in `/home/ethan/`

### Find out where you are

-   To find out where R is use `getwd()`

``` r
getwd()
```

-   "get working directory"
-   The "working directory" is where the program starts from

### Loading data

-   For data in the working directory just use the file name

``` r
shrub_data <- read.csv('shrub-dimensions-labeled.csv')
```

-   This is a relative path, because the file is in the working directory the only remaining piece is the name

-   One way to ensure that a file is in the working directory is to download it using `download.file()`

-   For data not in the working directory there are two options

    -   tell R where it is
    -   change the working directory to where it is

-   Changing the working directory is common

    -   Who here uses `setwd` regularly?
    -   Does that ever cause any issues?
        -   Working on a different computer?
        -   Working with someone else's files?

-   In general, using `setwd` means that your code will only work on a single computer

-   That's bad, so we want to have the working directory automatically and use relative paths

### Projects

-   The simplest way to do this is using RStudio Projects

-   (Posit Cloud) In fact we've already been using them

-   (Posit Cloud) Every time you click on an assignment or click `New Project` this creates a new project

-   Each project is a self-contained unit of work in a single folder/directory

-   Treat all locations as relative to that single directory

-   To do this in regular RStudio we use projects

-   `File` -\> `New Project` -\> `New Directory` -\> `New Project` -\> datacarp

-   Or use `Existing Directory` to choose an existing directory

-   Creates .Rproj file

    -   Isn’t project itself
    -   Contains project info
    -   Don’t change manually

-   Then place your data files in the project directory (or a subdirectory)

-   *copy shrub-dimensions-labeled.csv to a /home/user/datacarp/*

```{r eval = FALSE}
# OSX/Linux
data <- read.csv('shrub-dimensions-labeled.csv')
```

-   It is common to store data in a subdirectory
-   `New Folder` -\> data -\> `OK`
-   `File checkbox` -\> `More` -\> `Move` -\> data

```{r eval = FALSE}
# OSX/Linux
data <- read.csv('data/shrub-dimensions-labeled.csv')
```

-   Can switch between projects using `File` -\> `Recent Projects` or `Open Project`
-   Keeps track the state of RStudio when you last worked with that project

### Paths in R Scripts vs. RMarkdown Files

For reasons I haven't quite been able to nail down, file paths work a little bit different in R scripts and RMarkdown files in RProjects.

### Sharing code

-   This idea of projects as folders is also important for how we share code
-   Version control, which we'll have the opportunity to learn about later, works with the projects as folder structure
-   We can also zip a the project folder and send it to someone else and they can work with it
