---
title: "Week 6 Assignment"
author: "Ellen Bledsoe"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Week 6 Assignment

# Assignment Exercises

### Set-up

Load the packages we will need. You can either load all of them individually (`readr`, `dplyr`, `tidyr`, `ggplot2`) or load the `tidyverse` package.

```{r}
library(tidyverse)
```

### 1. Forest Area per Country (20 pts)

These data are downloaded from the WHO and contain the amount of forest (sq. km) per country. The first 3 rows of the file are metadata or empty, which we do not want. I've added the arguments `skip = 4` and `col_names = TRUE` to the `read_csv` function to deal with this.

```{r}
forest <- read_csv("forest_per_country.csv", skip = 4, col_names = TRUE)
```

a.  Currently, this data is in a wide format. We want to convert this to a longer format and make it tidy. Use the `pivot_longer` function to do so. Overwrite the `forest` dataframe so that it contains the long version of the data.

    Because the column names start with numbers, which R does not like, we need to put the column names either in backticks or quotation marks (e.g., "1990":"2022").

b.  Remove any rows that have `NA` in the forest area column using the `drop_na()` function.

c.  Let's remind ourselves how to plot the data. Make a scatterplot of the data with year on the x-axis and forest area on the y-axis. Make the points partially tranparent and the color "forestgreen." Add more descriptive axes labels and a theme.

    Add the following line of code to the end of your ggplot code so we can see the years along the x-axis: `theme(axis.text.x = element_text(angle = 45, vjust = 0.5))`.

```{r}
# a
forest <- forest %>% 
  pivot_longer(`1990`:`2022`, names_to = "Year", values_to = "ForestArea_sqkm")
forest

# b.
forest <- forest %>% 
  drop_na(ForestArea_sqkm)
forest
  
# c
ggplot(forest, aes(Year, ForestArea_sqkm)) +
  geom_point(alpha = 0.5, color = "forestgreen") +
  labs(x = "Year",
       y = "Forest Area (sq. km)") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5))
```

### 2. OECD Data (20 pts)

```{r}
oecd <- read_csv("oecd_annual_data.csv", skip = 4, col_names = TRUE)
head(oecd)
```

a.  fill
b.  pivot longer
c.  histogram of data, colored by OECD member

```{r}
#a
oecd <- oecd %>% 
  fill(OECD_member)

#b
oecd <- oecd %>% 
  pivot_longer(`2000`:`2022`, names_to = "Year", values_to = "MarineProtectedArea_sqkm")
```

### 3. Santa Cruz Rodents Data Cleaning (20 pts)

Start by reading in the rodent data from the Santa Cruz River, `capture_data.csv`.

```{r}
rodents <- read_csv("capture_data.csv")
```

Take a look at the data. You'll likely notice immediately that there are some issues to be fixed.

a.  Rename any column that needs to be renamed and save the output. This is the data frame we will use for the remainder of the questions.

b.  Next we need to fill in the missing values in the `Site` column.

c.  In the `Species` column, there are 2 different species that have question marks next to their names. Using the `replace` function inside of a `mutate` function, remove the question marks (e.g., `SIOC?` should become `SIOC` and `DIME?` should become `DIME`.

    (This is just for practice. In reality, we might want to create a code for unknown species or a column for unclear ID).

d.  If we look at the data classes for the columns, we can see that the column for tail length is character when it should be numeric. This usually indicates that there is a special character or letter somewhere in the column. As it turns out, in the last row, the value is `~15.5` instead of `15.5`. Use the `replace` function inside of `mutate` to convert that value to `15.5`.

e.  In both the Hair Sample and Position columns, there is a `?`. Use the `na_if` function inside a `mutate` function to convert those `?` to `NA` values.

```{r}
#a
rodents <- rodents %>% 
  rename(TrapID = `Trap ID`, Status = `Status (R/N)`, TotalWeight = `Total Weight`,
         BagWeight = `Bag weight`, AnimalWeight = `Animal Weight`, 
         HindfoodLength = `Hind foot length`, TailLength = `Tail length`,
         HairSample = `Hair sample (Y/N)`, Position = `Position (R/L)`)

#b 
rodents <- rodents %>% 
  fill(Site)

#c
rodents <- rodents %>% 
  mutate(Species = replace(Species, Species == "SIOC?", "SIOC"),
         Species = replace(Species, Species == "DIME?", "DIME"))

#d
rodents <- rodents %>% 
  mutate(TailLength = replace(TailLength, TailLength == "~15.5", "15.5"))

#e
rodents <- rodents %>% 
  mutate(HairSample = na_if(HairSample, "?"),
         Position = na_if(Position, "?"))

```

### 4. Santa Cruz Rodents Wrangling (20 pts)

Let's practice splitting and combining columns as well as pivoting the Santa Cruz rodent data.

a.  Use the `separate()` function to split the date column into 3 separate columns.

b.  Use `unite()` to bring them back together into 1 Date column

c.  Summarize the data so that we have a count of each species per site. Save this output as a new data frame (do not overwrite the rodents data frame)

d.  Convert the data from (c) from long format to wide format. Use an argument in the `pivot_wider` function to have all blank cells filled with 0 instead of `NA`.

```{r}
#a
rodents <- rodents %>% 
  separate(Date, into = c("Year", "Month", "Day"), sep = "-")
rodents

#b
rodents <- rodents %>% 
  unite("Date", Year:Day, sep = "-")
rodents

#c
sp_by_site <- rodents %>% 
  group_by(Site, Species) %>% 
  summarize(Count = n())
sp_by_site

#d
sp_by_site %>% 
  pivot_wider(names_from = Species,
              values_from = Count,
              values_fill = 0)
```

### 5. Mammals (20 points

```{r}
mammals <- data.frame(site = c(1,1,2,3,3,3), 
                      taxon = c('Suncus etruscus', 'Sorex cinereus', 
                                'Myotis nigricans', 'Notiosorex crawfordi', 
                                'Suncus etruscus', 'Myotis nigricans'),
                      density = c(6.2, 5.2, 11.0, 1.2, 9.4, 9.6),
                      mass = c(4.2, 5, 9.1, 8.6, 4.1, 8.7))
```

a.  separate
b.  pivot longer
c.  unite
d.  pivot wider

```{r}
mammals <- mammals %>% 
  separate(taxon, c("genus", "species"), sep = " ")

mammals <- mammals %>% 
  pivot_longer(density:mass, names_to = "measurement", values_to = "value")

mammals <- mammals %>% 
  unite("taxon", genus, species, sep = " ")

mammals <- mammals %>% 
  pivot_wider(names_from = measurement, values_from = value)
```
