---
title: "Assignment 9"
author: "Ellen Bledsoe"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Assignment Details

### Purpose

The goal of this assignment is to practice problem decomposition and some best practices in reproducibility .

### Task

Write R code to successfully answer each question below.

### Criteria for Success

-   Code is within the provided code chunks or new code chunks are created where necessary
-   Code chunks run without errors
-   Code chunks have brief comments indicating which code is answering which part of the question
-   Code will be assessed as follows:
    -   Produces the correct answer using the requested approach: 100%
    -   Generally uses the right approach, but a minor mistake results in an incorrect answer: 90%
    -   Attempts to solve the problem and makes some progress using the core concept, but returns the wrong answer and does not demonstrate comfort with the core concept: 50%
    -   Answer demonstrates a lack of understanding of the core concept: 0%
-   Any questions requiring written answers are answered with sufficient detail

### Due Date

March 25 at midnight MST

# Assignment Exercises

For many of the exercises in this week's assignment, we will actually be using a lot of the code that you have already written for Assignment 8. This time, however, all of your file paths will be different...

### 1. Set-Up (20 pts)

Now that we are working outside of Posit Cloud, we will need to first *install* our packages onto your computer before we can load them with the `library()` function.

We use the `install.packages()` function to download the package from the internet and create a local copy. Unlike with the `library()` function, the package name needs to be inside quotation marks.

Insert a code chunk and (a) install and (b) load `palmerpenguins`. Since we already installed the `tidyverse` during the lesson, you do not need to install it again. However, you do need to *load* it again.

Once you've installed `palmerpenguins`, comment out that line of code.

#### Note!

If you haven't already, make sure you are working in an RStudio Project! 

Your project should have sub-directories for raw data, clean data, output, docs, and scripts. The naming convention for these folders is up to you.

Make sure that this .Rmd file is in the appropriate sub-directory.

### 2. Portal Data Paths Review (20 points)

For this question, we are going to be using some of the code you've already written for Assignment 8, Question 2.

Click on the links below to download all 3 of the Portal files: surveys, species, and plots.

```{r}
# data
download.file("https://ndownloader.figshare.com/files/2292172",
  "surveys.csv")
download.file("https://ndownloader.figshare.com/files/3299474",
  "plots.csv")
download.file("https://ndownloader.figshare.com/files/3299483",
  "species.csv")
```


Then, move those three files from your Downloads folder into your equivalent of the raw data folder (yours might have a slightly different name). This assignment file should be in your scripts folder.

Now, let's begin to code!

a.  Load the 3 data frames (surveys, species, plots) into R using `read_csv()`. Make sure your paths are *relative*.
b.  Copy the answers from Week 8 Assignment, Questions 2d-f, into the code chunk below.
c.  Save the output of the code from 2d as a new data frame. Then, save that resulting data frame as a csv file in the clean data sub-directory using the `write_csv()` function.
d.  Save the ggplots from 2e and 2f into the outputs folder using the `ggsave()` function.

### 3. Palmer Penguins and Path Files (20 points)

Like in Question 2 above, we will be recreating Question 4 from Week 8's assignment but within our own R Project. As a reminder, this is the question that used the `palmerpenguins` data.

a.  The code chunk below uses the `download.file()` function to go to a specific URL and then download the data at that URL. The location where the file is downloaded to is set by the `destfile` argument. Modify the path in the `destfile` argument for all three species datasets so that they are downloaded directly into your raw data folder.

```{r}
# Adelie penguin data
download.file(url = "https://portal.edirepository.org/nis/dataviewer?packageid=knb-lter-pal.219.3&entityid=002f3893385f710df69eeebe893144ff",
              destfile = "adelie.csv")

# Gentoo penguin data
download.file(url = "https://portal.edirepository.org/nis/dataviewer?packageid=knb-lter-pal.220.3&entityid=e03b43c924f226486f2f0ab6709d2381",
              destfile = "gentoo.csv")

# Chinstrap penguin data 
download.file(url = "https://portal.edirepository.org/nis/dataviewer?packageid=knb-lter-pal.221.2&entityid=fe853aa8f7a59aa84cdd3197619ef462",
              destfile = "chinstrap.csv")
```

b.  Copy, modify, and then run the code that you wrote to combine the three above datasets and have the output match the `penguins` dataframe from the `palmerpenguins` dataframe.
c.  Run the `setdiff()` function to make sure that your code worked (it shouldn't have any issues, but it is good to check!)
d.  Save your version of the cleaned penguins data into your clean data folder.

### 4. Usings R Scripts (20 points)

Finally, we will be recreating Question 3 for Week 8's Assignment.

This time, however, you will not be doing so in this .Rmd file. Instead, I want you to make and use an R script!

a. Create a new R script or use the one we used in class--either is fine. Make sure it is in the correct sub-directory of your RStudio project. If you create a new script, I highly recommend saving it before you continue with this question (unsaved files can lead to wonky file path issues).
b. Ensure that the `mammal-size-data-clean` text file (download from D2L) is in the appropriate sub-directory in your RStudio Project.
c. Now, using the code your previously wrote, modify the code so that it runs in your R script file. 
d. Again, use the `ggsave()` function to save the figures from 3b, 3c, and 3d into your `output` (or similar) sub-directory.

### 5. Final Project Proposal (20 points)

In D2L, you will find a file with a few questions about what dataset you are planning to use. If you don't have one in mind yet, don't worry! I'll provide a list with some options for you to explore.

Download the file from D2L and place it in the documents folder of your RStudio Project.

## Turning in Your Assignment

To turn in your assignment, you will need to upload the entire folder that contains your .Rproj file and sub-directories. 

Usually, the easiest way to do this is to go to your Files application, navigate to the folder, and right click on the folder you want to compress. You should see the option to "Compress", "Compress to ZIP file", or something along those lines. Upload the ZIP file to D2L.
