---
title: 'Week 3: Combining Steps with Pipes'
author: "Ellen Bledsoe"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Combining Data Manipulations in dplyr

Introduction
Combine a series of data manipulation actions
Do each action in sequential order
Intermediate variables
Run a command
Store the output in a variable
Use that variable later in the code
Repeat

Obtain the data for only DS, sorted by year, with only the year and and weight columns
```{r}
ds_data <- filter(surveys, species_id == "DS", !is.na(weight))
ds_data_by_year <- arrange(ds_data, year)
ds_weight_by_year <- select(ds_data_by_year, year, weight)
```

Do Portal Data Manipulation (Exercise 3)

Pipes
Intermediate variables can get cumbersome if their are lots of steps.
|> (“pipe”) takes the output of one command and passes it as input to the next command
Want to take the mean of a vector
Normally we would run the mean function with the vector as the input:
```{r}
x = c(1, 2, 3)
mean(x)
```

Instead we could pipe the vector into the function
```{r}
x |> mean()
```

So x becomes the first argument in mean
If we want to add other arguments they get added to the function call
```{r}
x = c(1, 2, 3, NA)
mean(x, na.rm = TRUE)
x |> mean(na.rm = TRUE)
```

Let's try this with some of the `dplyr` functions we have learned.
```{r}
surveys |>
  filter(species_id == "DS", !is.na(weight))
ds_weight_by_year <- surveys |>
  filter(species_id == "DS", !is.na(weight)) |>
  arrange(year) |>
  select(year, weight)
```

Try your hand at the Portal Data Manipulation Pipes (Exercise 4).

The magrittr pipe
You will also see another type of pipe character %>%
This is the original pipe in R and you had to load the magrittr package to use it (this gets loaded automatically by dplyr)
Either pipe is fine for this class
|> will work everywhere as long as you have a new enough version of R
magrittr has some fancier functionality that may be useful in some cases
Keyboard Shortcut
Shortcut: Ctrl-Shift-m
You can change this to give the base R pipe
Tools -> Global Options -> Code -> Use native pipe operator
Pipe to other arguments
To pipe the result of a line to something other than the first argument use the placehold _
This only works for named arguments
Let’s fit a linear model at the end of our dplyr pipeline
lm takes a formula as the first argument tells it what columns to use for the response and driver variables
The second argument tells it where the data is
It needs to be named for the place holder to work
```{r}
surveys |>
  filter(species_id == "DS", !is.na(weight)) |>
  arrange(year) |>
  select(year, weight) |>
  lm(weight ~ year, data = _)
```

