---
title: 'Assignment 11: Conditionals'
author: "Ellen Bledsoe"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Assignment


### 1. Choice Operators (25 pts)

Create the following variables.

```{r}
w <- 10.2
x <- 1.3
y <- 2.8
z <- 17.5
colors <- c("red", "blue", "green")
masses <- c(45.2, 36.1, 27.8, 81.6, 42.4)
dna1 <- "attattaggaccaca"
dna2 <- "attattaggaacaca"
```

Use them to print whether or not the following statements are `TRUE` or `FALSE`.

a. `w` is greater than 10
b. `"green"` is in `colors`
c. `x` is greater than `y`
d. Each value in `masses` is greater than 40.
e. `dna1` is the same as `dna2`
f. `dna1` is not the same as `dna2`
g. `w` is greater than `x`, or `y` is greater than `z`
h. `x` times `w` is between 13.2 and 13.5
i. Each mass in `masses` is between 30 and 50.

```{r}
### Code solution for Choice Operators

print("1a")
# a.  `w` is greater than 10
w > 10

print("1b")
# b.  `w` + `x` is less than 15
w + x < 15

print("1c")
# c.  `x` is greater than `y`
x > y

print("1d")
# d. Each value in `masses` is greater than 50.
masses > 40

print("1e")
# f.  `dna1` is the same as `dna2`
dna1 == dna2

print("1f")
# g.  `dna1` is not the same as `dna2`
dna1 != dna2

print("1g")
# h. `w` is greater than `x`, or `y` is greater than `z`
w > x | y > z

print("1h")
# i.  `x` times `w` is between 13.2 and 13.5
13.2 < x * w & x * w < 13.5  # | (a pipe) will return TRUE for all values.

print("1i")
# j. Each mass in `masses` is between 30 and 50.
masses > 30 & masses < 50
```


### 2. If Statements (25 points)

a. Complete the following `if` statement so that if `age_class` is equal to "sapling" it sets `y <- 10`.

```{r}
age_class = "sapling"
if (){
  
}
y
```

b. Complete the following `if` statement so that if `age_class` is equal to
   "sapling" it sets `y <- 10` and if `age_class` is equal to "seedling" it
   sets `y <- 5`.

```{r}
age_class = "seedling"
if (){
  
}
y
```

c. Complete the following `if` statement so that if `age_class` is equal to
   "sapling" it sets `y <- 10` and if `age_class` is equal to "seedling" it
   sets `y <- 5` and if `age_class` is something else then it sets the value of
   `y <- 0`.

```{r}
age_class = "adult"
if (){
  
}
y
```

d. Convert your conditional statement from (c) into a function that takes
   `age_class` as an argument and returns `y`. Call this function 5 times, once
   with each of the following values for `age_class`: "sapling", "seedling",
   "adult", "mature", "established".

```{r}
print("2a")
# a. Complete (i.e., copy into your code and them modify) the following `if`
#   statement so that if `age_class` is equal to "sapling" it sets `y <- 10`.

age_class = "sapling"
if (age_class == "sapling"){
  y <-10
}
y

print("2b")
# b. Complete the following `if` statement so that if `age_class` is equal to
#   "sapling" it sets `y <- 10` and if `age_class` is equal to "seedling" it
#   sets `y <- 5`.

age_class = "seedling"
if (age_class == "sapling"){
  y <-10
} else if (age_class == "seedling"){
  y <- 5
}
y


print("2c")
# c. Complete the following `if` statement so that if `age_class` is equal to
#   "sapling" it sets `y <- 10` and if `age_class` is equal to "seedling" it
#   sets `y <- 5` and if `age_class` is something else then it sets the value of
#   `y <- 0`.

age_class = "adult"
if (age_class == "sapling"){
  y <-10
} else if (age_class == "seedling") {
  y <- 5
} else {
  y <- 0
}
y

print("2d")
# d. Convert your conditional statement from (c) into a function that takes
#   `age_class` as an argument and returns `y`. Call this function 5 times, once
#   with each of the following values for `age_class`: "sapling", "seedling",
#   "adult", "mature", "established".

get_y_from_age_class <- function(age_class){
  if (age_class == "sapling"){
    y <-10
  } else if (age_class == "seedling") {
    y <- 5
  } else {
    y <- 0
  }
  return(y)
}

get_y_from_age_class("sapling")
get_y_from_age_class("seedling")
get_y_from_age_class("adult")
get_y_from_age_class("mature")
get_y_from_age_class("established")
```


### 3. If Statements in Functions (25 points)

a. Write a function named `double_if_small` that takes a number as input and returns the number multiplied by 2 if the input is less than 26 and returns just the number (not multiplied by two) if the input is greater than or equal to 26. Call the function with `10` as the input.
b. Call the function from (a) with `30` as the input.
c. Write a function called `prediction` that takes a single argument `x`. If `x` is both greater than 0 and less than 15 then return `y = 6 + 0.8 * x`. If `x` is both greater than 15 and less than 30 then return `y = 5 + 0.75 * x`. In all other cases return `y = NA`. Call the function with `5` as the input.
d. Call the function from (c) with `26` as the input.
e. Call the function from (c) with `-2` as the input.

```{r}
# Solution to If Statements in Functions exercise

print("3a")
double_if_small <- function(number){
  if (number < 26){
    output = number * 2
  } else {
    output = number
  }
  return(output)
}

double_if_small(10)

print("3b")

double_if_small(30)

print("3c")

prediction <- function(x){
  if (x > 0 & x < 15){
      y = 6 + 0.8 * x
  } else if (x > 15 & x < 30) {
      y = 5 + 0.75 * x
  } else {
      y = NA
  }
  return (y)
}

prediction(5)

print("3d")

prediction(26)

print("3e")

prediction(-2)
```


### 4. Size Estimates by Name (25 points)

This is a follow up to "Use and Modify" from Assignment 10.

To make it even easier to work with your dinosaur size estimation functions you
decide to create a function that lets you specify which dinosaur group you need
to estimate the size of by name and then have the function automatically choose
the right parameters.

Remember the general form of the equation is:

> mass <- a * length ^ b

Create a new function `get_mass_from_length_by_name()` that takes two arguments,
the `length` and the name of the dinosaur group. Inside this function use
`if`/`else if`/`else` statements to check to see if the name is one of the
following values and if so use the associated `a` and `b` values to estimate the
species mass using these equations:

* *Stegosauria*:  `mass = 10.95 * length ^ 2.64` ([Seebacher 2001](http://www.jstor.org/stable/4524171))
* *Theropoda*:  `mass = 0.73 * length ^ 3.63` ([Seebacher 2001](http://www.jstor.org/stable/4524171))
* *Sauropoda*:  `mass = 214.44 * length ^ 1.46` ([Seebacher 2001](http://www.jstor.org/stable/4524171))

If the name is not any of these values the function should return `NA`.

Run the function for:
a. A *Stegosauria* that is 10 meters long.
b. A *Theropoda* that is 8 meters long.
c. A *Sauropoda* that is 12 meters long.
d. A *Ankylosauria* that is 13 meters long.

```{r}
get_mass_from_length <- function(a, b, length){
  mass = a * length ** b
  return (mass)
}

get_mass_from_length_by_name <- function(length, name){
  if (name == "Stegosauria"){
    if (length > 8){
      a = 10.95
      b = 2.64
    } else {
      a = 8.5
      b = 2.8
    }
  }
  else if (name == "Theropoda"){
    a = 0.73
    b = 3.63
  }
  else if (name == "Sauropoda"){
    a = 214.44
    b = 1.46
  }
  else {
    return(NA)
  }
  get_mass_from_length(a, b, length)
}

get_mass_from_length_by_name(10, "Stegosauria")
get_mass_from_length_by_name(8, "Theropoda")
get_mass_from_length_by_name(12, "Sauropoda")
get_mass_from_length_by_name(13, 'Ankylosauria')
```

