---
title: "Week 4 Assignment"
author: "Ellen Bledsoe"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Assignment Details

### Purpose

The goal of this assignment is to work with data aggregation and joining data frames together using `dplyr` functions.

### Task

Write R code to successfully answer each question below.

### Criteria for Success

-   Code is within the provided code chunks or new code chunks are created where necessary
-   Code chunks run without errors
-   Code chunks have brief comments indicating which code is answering which part of the question
-   Code will be assessed as follows:
    -   Produces the correct answer using the requested approach: 100%
    -   Generally uses the right approach, but a minor mistake results in an incorrect answer: 90%
    -   Attempts to solve the problem and makes some progress using the core concept, but returns the wrong answer and does not demonstrate comfort with the core concept: 50%
    -   Answer demonstrates a lack of understanding of the core concept: 0%
-   Any questions requiring written answers are answered with sufficient detail

### Due Date

Feb 12 at midnight MST

# Assignment Exercises

### 1. Set-Up (10 points)

Load the `readr` and `dplyr` packages.

```{r}
library(readr)
library(dplyr)
```

Read in the following data sets using `read_csv()`:

-   `surveys.csv`
-   `species.csv`
-   `plots.csv`

```{r}
surveys <- read_csv("surveys.csv")
species <- read_csv("species.csv")
plots <- read_csv("plots.csv")
```

### 2. Portal Data Aggregation (10 pts)

Using the `surveys` data frame, complete the following:

a.  Use the `group_by()` and `summarize()` functions to get a count of the number of individuals in each species ID.
b.  Use the `group_by()` and `summarize()` functions to get a count of the number of individuals in each species ID in each year.
c.  Use the `filter()`, `group_by()`, and `summarize()` functions to get the mean mass of species `DO` in each year.

```{r}
# 1. Use the group_by() and summarize() functions to get a count of the number of individuals in each species ID.

surveys %>% 
  group_by(species_id) %>% 
  summarize(count = n())

# 2. Use the group_by() and summarize() functions to get a count of the number of individuals in each species ID in each year.

surveys %>% 
  group_by(species_id, year) %>% 
  summarize(count = n())

# 3. Use the filter(), group_by(), and summarize() functions to get the mean mass of species DO in each year.
 
 surveys %>%
  filter(species_id == "DO") %>%
  group_by(year) %>%
  summarize(avg_mass = mean(weight, na.rm = TRUE))

```

### 3. Shrub Volume Aggregation (15 pts)

This is a follow-up to Shrub Volume Data Basics (from a previous assignment).

Dr. Morales wants some summary data of the plants at her sites and for her experiments. The following code calculates the average height of a plant at each site:

```{r}
shrub_dims <- read_csv('shrub-volume-data.csv')
shrub_dims %>% 
  group_by(experiment) %>% 
  summarize(avg_height = mean(height, na.rm = TRUE),
            max_height = max(height, na.rm = TRUE))
```

Modify the code to calculate and print the average height of a plant in *each experiment*.

Add a line of code to use `max()` to determine the maximum height of a plant at each experiment.

Also, remember to modify the code so that there are no NAs produced in the final output.

### 4. Portal Data Joins (15 pts)

Using the Portal data sets, do the following:

a.  Use `inner_join()` to create a table that contains the information from both the surveys table and the species table.
b.  Use `inner_join()` twice to create a table that contains the information from all three tables.
c.  Use `inner_join()` and `filter()` to get a data frame with the information from the surveys and plots tables where the plot_type is Control.

```{r}
print("4a")

inner_join(surveys, species, by = "species_id")

print("4b")
inner_join(surveys, species, by = "species_id") %>% 
inner_join(., plots, by = "plot_id")

print("4c")

surveys %>% 
  inner_join(plots, by = "plot_id") %>% 
  filter(plot_type == "Control")
```

### 5. Portal Data `dplyr` Review (25 pts)

We want to do an analysis comparing the size of individuals on the Control plots to the Long-term Krat Exclosures.

a. Create a data frame with the `year`, `genus`, `species`, `weight` and `plot_type` for all cases where the plot type is either "Control" or "Long-term Krat Exclosure." Only include cases where Taxa is "Rodent." Remove any records where the weight is missing.
b. Now, use the "split, apply, combine" approach to calculate the minimum, maximum, and average weight per plot type per year.

```{r}
print("5a")
surveys %>% 
  inner_join(species, by = "species_id") %>% 
  inner_join(plots, by = "plot_id") %>% 
  filter(plot_type == "Control" | plot_type == "Long-term Krat Exclosure") %>% 
  filter(taxa == "Rodent") %>% 
  select('year', 'genus', 'species', 'weight', 'plot_type') %>% 
  filter(!is.na(weight))

print("5b")
surveys %>% 
  inner_join(species, by = "species_id") %>% 
  inner_join(plots, by = "plot_id") %>% 
  filter(plot_type == "Control" | plot_type == "Long-term Krat Exclosure",
         taxa == "Rodent", 
         !is.na(weight)) %>% 
  select('year', 'genus', 'species', 'weight', 'plot_type') %>% 
  group_by(year, plot_type) |> 
  summarize(min_weight = min(weight),
            max_weight = max(weight),
            mean_weight = mean(weight))

```

### 6. Shrub Volume Bind (10 pts)

First, run the following code chunk to produce a data frame with additional data related to the shrub volume data (`shrub_dims`) from Question 3.

```{r}
new_data <- data.frame(respiratory_rate = c(2.2, 4.0, 6.1, 2.3, 4.1, 6.2, 1.8, 3.5, 5.7, 1.9, 3.5, 5.8, 2.0, 4.7, 6.2),
                       average_temp_C = c(15.1, 20.2, 24.7, 15.2, 22.0, 25.1, 14.2, 19.0, 23.6, 14.9, 20.3, 24.1, 19.2, 22.7, 25.0))

```

Take a look at the new dataframe that has just been produced. Should this data be bound to the shrub volume data by `bind_rows()` or `bind_cols()`? In your own words, explain how you know.

*Answer:*

Based on your answer above, bind the `shrub_dims` and `new_data` data frames together.

```{r}
bind_cols(shrub_dims, new_data)
```

### 7. Shrub Volume Join (15 pts)

This is a follow-up to Question 3, Shrub Volume Aggregation.

In addition to the main data table on shrub dimensions, Dr. Morales has two additional data tables. The first describes the manipulation for each experiment. The second provides information about the different sites. Run the following code chunk to bring them into your environment.

```{r}
experiments <- read_csv("https://datacarpentry.org/semester-biology/data/shrub-volume-experiments.csv")
sites <- read_csv("https://datacarpentry.org/semester-biology/data/shrub-volume-sites.csv")
```

a. Combine the `experiments` data frame with the shrub dimensions data to add a manipulation column to the shrub data.

b. Next, combine the `sites` data frame with both the data on shrub dimensions and the data on experiments to produce a single data frame that contains all of the data. Save this data frame as `shrub_data`.

```{r}
# grading note; if someone saved over the `shrubs_dim` dataframe in question 3, then they might have a different answer

# 1. Import the experiments data and then use `inner_join` to combine
print("7a")
inner_join(shrub_dims, experiments, by = "experiment")

# 2. Import the sites data and the combine it with both the data on shrub
# dimensions and the data on experiments to produce a single data frame
# that contains all of the data.
print("7b")
shrub_dims_experiments <- inner_join(shrub_dims, experiments, by = "experiment")
shrub_data <- inner_join(shrub_dims_experiments, sites, by = "site")
shrub_data
```

