---
title: "Week 14 Assignment"
output: pdf_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Assignment

### Purpose

The goal of this assignment is to practice using parallelization.

### Task

Write R code to successfully answer each question below.

### Criteria for Success

-   Code is within the provided code chunks or new code chunks are created where necessary
-   Code chunks run without errors
-   Code chunks have brief comments indicating which code is answering which part of the question
-   Code will be assessed as follows:
    -   Produces the correct answer using the requested approach: 100%
    -   Generally uses the right approach, but a minor mistake results in an incorrect answer: 90%
    -   Attempts to solve the problem and makes some progress using the core concept, but returns the wrong answer and does not demonstrate comfort with the core concept: 50%
    -   Answer demonstrates a lack of understanding of the core concept: 0%
-   Any questions requiring written answers are answered with sufficient detail

### Due Date

April 29 at midnight MST

# Assignment Exercises

### 1. Set Up (10 points)

We will use a number of packages to complete this week's assignments. Load in the following packages: `tidyverse`, `foreach`, and `doParallel`.

### 2. Use `apply` functions (30 points)

```{r}
seq_list <- as.list(seq(1, 500, by = 5))
```

Use the `head` function to look at the `seq_list` object. You will notice that the output looks different from the vectors and dataframes that we are used to looking at. That is because it is a list, a collection of data objects. But don't worry! The code for `lapply` and `mclapply` is the same as in the lesson.

```{r}
output <- lapply(seq_list, sqrt)
output <- do.call(rbind, output)
```


```{r}
output2 <- mclapply(seq_list, sqrt, mc.cores = nCores)
output2 <- do.call(rbind, output2)
```


### 3. Use `foreach` (30 points)

```{r}
source("dino_allometry_fxn.R")

dino_data <- read_csv("../Week12_Iteration/data_raw/dinosaur_lengths.csv")
```


Use `.combine = c` in place of the `.combine = rbind`.

```{r}
results2 <- foreach::foreach(i = 1:nrow(dino_data), .combine = c) %do% {
  mass_from_length(dino_data$lengths[i],
                   dino_data$species[i])
}
```

Use parallel processing
```{r}
nCores <- parallel::detectCores() / 2

doParallel::registerDoParallel()

results3 <- foreach::foreach(i = 1:nrow(dino_data), .combine = c) %dopar% {
  mass_from_length(dino_data$lengths[i],
                   dino_data$species[i])
}

doParallel::stopImplicitCluster()
```

d. Go back through all of your code for question 3 and explicitly call all of the functions that you use. The ones you've written yourself do not need to be explicilty called.

### 4. Forest Change through Time (30 points)

```{r}
forest <- read_csv("https://raw.githubusercontent.com/bleds22e/DataWrangling/main/Week06_TidyData/forest_per_country.csv?token=GHSAT0AAAAAACRA45BL2D5ZS76ZJKYVOCDSZRHXYXA", 
                   skip = 4, col_names = TRUE) %>% 
  select(-`2022`)
```
(a) Write a for loop that calculates the change in forest per country from 1990 to 2021 by subtracting the 1990 value from the 2021 value. 

The output should be stored in a vector called `change_in_forest`.

```{r}
change_in_forest <- vector(mode = "numeric", length = nrow(forest))

for (i in 1:nrow(forest)) {
  change_in_forest[i] <- forest$`2021`[i] - forest$`1990`[i]
}
```

(b) Perform the same task as in (a) using the `foreach` function and the `%do%` operator. Use the `.combine = rbind` argument to save the output as a dataframe.

```{r}
change_in_forest2 <- foreach(i = 1:nrow(forest), .combine = c) %do% {
  area <- forest$`2021`[i] - forest$`1990`[i]
}

```

(c) Do the same as in (b) but use parallel processing and the `%dopar%` operator.

```{r}
registerDoParallel(nCores)

change_in_forest3 <- foreach(i = 1:nrow(forest), .combine = c) %dopar% {
  area <- forest$`2021`[i] - forest$`1990`[i]
}

stopImplicitCluster()
```

