---
title: "Prep"
author: "Ellen Bledsoe"
date: "`r Sys.Date()`"
output: pdf_document
---


Edit Macroplots data
```{r}
macroplots <- read_tsv("http://datacarpentry.org/semester-biology/data/Macroplot_data_Rev.txt")

# put NAs in SpCode column to use `fill()` function
macroplots <- macroplots |> 
  mutate(SpCode = if_else(SpCode == lag(SpCode), NA, SpCode))

# replace first row in SpCode with the original species code
macroplots[1, "SpCode"] <- "Acaccaes"

# change column names to use `rename()` function
colnames(macroplots) <- c("Plot ID", "Sp Code", "TreeGirth1", "TreeGirth2", "TreeGirth3", "TreeGirth4", "TreeGirth5")

# save output to data folder for lesson
write_tsv(macroplots, "data/macroplot_data.txt")
```





```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Number of species in `trees.txt`

```{r}
# 6 + NA
trees %>% 
  distinct(SPECIES)
```

```{r}
# remove NA species
trees <- trees %>% 
  filter(!is.na(SPECIES))
```

Demonstrate separate and unite with species column
```{r}
# separate

genus_sp <- trees %>% 
  separate(SPECIES, into = c("GENUS", "SPECIES"), sep = "_")


# unite
genus_sp %>% 
  unite("SPECIES", GENUS:SPECIES, sep = " ")

genus_sp <- genus_sp %>% 
  unite("SPECIES", GENUS:SPECIES, sep = " ", na.rm = TRUE)

```

Convert to pivot_wider (species by site)

```{r}
sp_by_site <- genus_sp %>% 
  group_by(TREATMENT, SPECIES) %>% 
  summarise(ABUND = n())

sp_by_site %>% 
  pivot_wider(names_from = TREATMENT, values_from = ABUND )
```
## Practice with santa cruz rodents

```{r}
rodents <- read_csv("capture_data.csv")
plants <- read_csv("microsite_raw.csv")
```

Trying to select with spaces
```{r}
rodents %>% select(Date, Site, Trap ID)
```

Use backticks to deal with names with spaces

```{r}
rodents %>% select(Date, Site, `Trap ID`)
```
But that gets annoying after a while.

We can use a tidyverse function called `rename` to address this issue

```{r}
rodents %>% 
  rename(TrapID = `Trap ID`) %>% 
  select(Date, Site, TrapID)
```

Create day, month, year column
```{r}
yr_mo_day <- rodents %>% 
  separate(Date, into = c("Year", "Month", "Day"), sep = "-")
```

Reunite dates
```{r}
dates <- yr_mo_day %>% 
  unite("Date", Year:Day, sep = "-") %>% 
  mutate(Date = as_date(Date))
```

### Plant Data

```{r}
head(plants)
plants <- plants %>% 
  rename(VegType = `Type of Vegetation`) %>% 
  separate(VegType, into = c("Veg1", "Veg2"), sep = "/")

# pivot longer
plants <- plants %>% 
  pivot_longer(Veg1:Veg2,
               names_to = "Veg",
               values_to = "VegetationType")

plants <- plants %>% 
  filter(!is.na(VegetationType)) %>% 
  select(-Veg)
```

## Good Links
https://biol355.github.io/Labs/axoltl_lab.html

https://biol355.github.io/Lectures/13_data_tidy.html

https://www.rforecology.com/post/reshaping-data-in-r/
----
