---
title: "Week 6 Assignment"
author: "Ellen Bledsoe"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Week 6 Assignment

### Purpose

The goal of this assignment is to practice converting messy data into tidy data.

### Task

Write R code to successfully answer each question below.

### Criteria for Success

-   Code is within the provided code chunks or new code chunks are created where necessary
-   Code chunks run without errors
-   Code chunks have brief comments indicating which code is answering which part of the question
-   Code will be assessed as follows:
    -   Produces the correct answer using the requested approach: 100%
    -   Generally uses the right approach, but a minor mistake results in an incorrect answer: 90%
    -   Attempts to solve the problem and makes some progress using the core concept, but returns the wrong answer and does not demonstrate comfort with the core concept: 50%
    -   Answer demonstrates a lack of understanding of the core concept: 0%
-   Any questions requiring written answers are answered with sufficient detail

### Due Date

March 4 at 3:30 pm

# Assignment Exercises

### Set-up

Load the packages we will need. You can either load all of them individually (`readr`, `dplyr`, `tidyr`, `ggplot2`) or load the `tidyverse` package.

```{r}
library(tidyverse)
```

### 1. Forest Area per Country (15 pts)

The data in `forest_per_country.csv` are downloaded from the WHO and contain the amount of forest (sq. km) per country. 

First, click on the .csv file in the "Files" tab and select "View File." You might notice that the top of the document looks a bit odd.

The first 3 rows of the file are metadata or empty, which we do not want. I've added the arguments `skip = 4` and `col_names = TRUE` to the `read_csv` function to deal with this.

```{r}
forest <- read_csv("forest_per_country.csv", skip = 4, col_names = TRUE)
```

a.  Currently, this data is in a wide format. We want to convert this to a longer format and make it tidy. Use the `pivot_longer` function to do so. Overwrite the `forest` dataframe so that it contains the long version of the data.

    Because the column names start with numbers, which R does not like, we need to put the column names either in backticks or quotation marks (e.g., "1990":"2022").

b.  Remove any rows that have `NA` in the forest area column using the `drop_na()` function.

c.  Let's remind ourselves how to plot the data. Make a scatterplot of the data with year on the x-axis and forest area on the y-axis. Make the points partially tranparent and the color "forestgreen." Add more descriptive axes labels and a theme.

    Add the following line of code to the end of your ggplot code so we can see the years along the x-axis: `theme(axis.text.x = element_text(angle = 45, vjust = 0.5))`.

```{r}
# a
forest <- forest %>% 
  pivot_longer(`1990`:`2022`, names_to = "Year", values_to = "ForestArea_sqkm")
forest

# b.
forest <- forest %>% 
  drop_na(ForestArea_sqkm)
forest
  
# c
ggplot(forest, aes(Year, ForestArea_sqkm)) +
  geom_point(alpha = 0.5, color = "forestgreen") +
  labs(x = "Year",
       y = "Forest Area (sq. km)") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5))
```

### 2. OECD Data (10 pts)

The `oecd_annual_data.csv` contains data from the Organisation for Economic Co-operation and Development (OECD) about various global fishing economies and sustainability. This dataset has the area of protected marine reserves.

a. Like the forest data from Question 1, this data has a few rows of metadata at the top of the document that we need to skip. Use the same arguments as we did in Question 1 (same values, as well) when reading in the file.

```{r}
oecd <- read_csv("oecd_annual_data.csv", skip = 4, col_names = TRUE)
head(oecd)
```

b.  Use the `fill()` function to fill in the missing values in the first column. Save the output as an object to be used in (c).

c.  Using the data frame you created in (b), put the data in a tidy format. You'll need to use the same trick with the year column names as you did in 1a. Save the output as an object.

d.  Using the data frame you created in (c), only keep the rows for non-OECD members. Then, wrangle the data so that it is back in wide format. This time, however, each row should represent a year and each country will have its own column. Your final data frame should only have a year column and each country's column.


```{r}
#b
oecd <- oecd %>% 
  fill(OECD_member)
oecd

#c
oecd <- oecd %>% 
  pivot_longer(`2000`:`2022`, names_to = "Year", values_to = "MarineProtectedArea_sqkm")
oecd

#d
oecd |> 
  filter(OECD_member == "Non-OECD Economies") |> 
  pivot_wider(names_from = Country, values_from = MarineProtectedArea_sqkm) |> 
  select(-OECD_member)
```

### 3. Santa Cruz Rodents Data Cleaning (20 pts)

Start by reading in the rodent data from the Santa Cruz River, `capture_data.csv`.

```{r}
rodents <- read_csv("capture_data.csv")
```

Take a look at the data. You'll likely notice immediately that there are some issues to be fixed.

For this question, there isn't really a good way for me to show you the output in the answer key; you'll want to take a look at the data frame in R and/or make good use of the filter function to make sure the issue got fixed.

a.  Rename any column that needs to be renamed and save the output. This is the data frame we will use for the remainder of the questions, and you'll want to continue overwriting the output to use in each subsequent step.

b.  Next we need to fill in the missing values in the `Site` column.

c.  In the `Species` column, there are 2 different species that have question marks next to their names. Using the `replace` function inside of a `mutate` function, remove the question marks (e.g., `SIOC?` should become `SIOC` and `DIME?` should become `DIME`.

    (This is just for practice. In reality, we might want to create a code for unknown species or a column for unclear ID).

d.  If we look at the data classes for the columns, we can see that the column for tail length is character when it should be numeric. This usually indicates that there is a special character or letter somewhere in the column. As it turns out, in the last row, the value is `~15.5` instead of `15.5`. Use the `replace` function inside of `mutate` to convert that value to `15.5`.

e.  In both the "Hair Sample" and "Position" columns, there is a `?`. Use the `na_if` function inside a `mutate` function to convert those `?` to `NA` values.

f.  Let's practice plotting again! Create a series of histograms (subplots within one larger plot) of the rodents' total weights, one histogram per species. Have the color of the histograms be determined by the site. Each histogram should be partially transparent, overlapping (not stacked), and the width of the bins should be 10. Finally, edit the axes labels and apply a theme.

```{r}
#a
rodents <- rodents %>% 
  rename(TrapID = `Trap ID`, Status = `Status (R/N)`, TotalWeight = `Total Weight`,
         BagWeight = `Bag weight`, AnimalWeight = `Animal Weight`, 
         HindfoodLength = `Hind foot length`, TailLength = `Tail length`,
         HairSample = `Hair sample (Y/N)`, Position = `Position (R/L)`)

#b 
rodents <- rodents %>% 
  fill(Site)

#c
rodents <- rodents %>% 
  mutate(Species = replace(Species, Species == "SIOC?", "SIOC"),
         Species = replace(Species, Species == "DIME?", "DIME"))

#d
rodents <- rodents %>% 
  mutate(TailLength = replace(TailLength, TailLength == "~15.5", "15.5"))

#e
rodents <- rodents %>% 
  mutate(HairSample = na_if(HairSample, "?"),
         Position = na_if(Position, "?"))

#f
ggplot(rodents, aes(TotalWeight, fill = Site)) +
  geom_histogram(alpha = 0.5, position = "identity", binwidth = 10) +
  facet_wrap(~ Species) +
  labs(x = "Weight (g)", y = "Frequency") +
  theme_classic()
```

### 4. Remembering Joins (15 pts)

Let's remind ourselves about joins from Week 4.

Read in the vegetation data that goes along with the Santa Cruz rodent data. The .csv file is called `microsite_grouped_veg.csv`

```{r}
veg <- read_csv("microsite_grouped_veg.csv")

#a
veg <- veg %>% 
  rename(RecordID = `...1`, TrapID = `Trap Location`, DistancetoVeg_m = `Distance to Vegetation (m)`,
         VegetationType = `Type of Vegetation`, PercentCover = `Percent Veg Cover`, DistancetoWater_m = `Distance to Water (m)`)

#b
veg <- veg %>% 
  select(Site, TrapID, Grouped_Veg)

#c
inner_join(rodents, veg)
```

a.  Rename the columns that should be renamed. Use a consistent structure (and make the Site and Trap Location column names match those from the rodents data frame in Question 3). Save the output.

b.  Using the output from (a), select the Site, Trap Location and Grouped Veg columns and save those as a new data frame. Save the output.

c.  Using the output from (b) and the final rodent data frame from Question 3, use an `inner_join()` to join those two data frames.

d.  In your own words (~2-3 sentences), explain how the inner join in (c) worked.

### 5. Santa Cruz Rodents Wrangling (20 pts)

Let's practice splitting and combining columns as well as pivoting the Santa Cruz rodent data (from Question 3).

a.  Use the `separate()` function to split the date column into three separate columns. Save your output.

b.  Using the output from (a), rejoin the three date columns back together into one column.

c.  Summarize the data so that we have a count of each species per site. Save this output as a new data frame (do not overwrite the rodents data frame)

d.  Convert the data from (c) from long format to wide format. Use an argument in the `pivot_wider` function to have all blank cells filled with 0 instead of `NA`.


```{r}
#a
rodents <- rodents %>% 
  separate(Date, into = c("Year", "Month", "Day"), sep = "-")
rodents

#b
rodents <- rodents %>% 
  unite("Date", Year:Day, sep = "-")
rodents

#c
sp_by_site <- rodents %>% 
  group_by(Site, Species) %>% 
  summarize(Count = n())
sp_by_site

#d
sp_by_site %>% 
  pivot_wider(names_from = Species,
              values_from = Count,
              values_fill = 0)
```

### 6. Mammals (20 pts)

The code chunk below has some made-up mammal data. Run the code chunk below to complete Question 6.

```{r}
mammals <- data.frame(site = c(1,1,2,3,3,3), 
                      taxon = c('Suncus etruscus', 'Sorex cinereus', 
                                'Myotis nigricans', 'Notiosorex crawfordi', 
                                'Suncus etruscus', 'Myotis nigricans'),
                      density = c(6.2, 5.2, 11.0, 1.2, 9.4, 9.6),
                      avg_mass = c(4.2, 5, 9.1, 8.6, 4.1, 8.7))
```

Like with many of the questions before, save your output from each task to use in the next task. You can overwrite the `mammals` object each time if you would like to.

a.  Use the `separte()` function to create columns for the genus and species (from the taxon column).

b.  Use `pivot_longer` so that `density` and `mass` end up in one column and the values end up in another column.

c.  Even though the data from (b) is longer, it isn't tidier. Explain why not.

d.  Use the `unite()` function to bring the genus and species column back together as one column with whatever separator you choose.

e.  Use `pivot_wider()` to bring the data frame back to it's original state.

```{r}
mammals <- mammals %>% 
  separate(taxon, c("genus", "species"), sep = " ")

mammals <- mammals %>% 
  pivot_longer(density:avg_mass, names_to = "measurement", values_to = "value")

mammals <- mammals %>% 
  unite("taxon", genus, species, sep = " ")

mammals <- mammals %>% 
  pivot_wider(names_from = measurement, values_from = value)
```
