---
title: "Week 6 Assignment"
author: "Ellen Bledsoe"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Week 6 Assignment

# Assignment Exercises

### Set-up

Load the packages we will need. You can either load all of them individually (`readr`, `dplyr`, `tidyr`, `ggplot2`) or load the `tidyverse` package.

```{r message=FALSE}
library(tidyverse)
```

### 1. Forest Area per Country (15 pts)


```{r message=FALSE}
forest <- read_csv("forest_per_country.csv", skip = 4, col_names = TRUE)
```


```{r}
# a
forest <- forest %>% 
  pivot_longer(`1990`:`2022`, names_to = "Year", values_to = "ForestArea_sqkm")
forest

# b.
forest <- forest %>% 
  drop_na(ForestArea_sqkm)
forest
  
# c
ggplot(forest, aes(Year, ForestArea_sqkm)) +
  geom_point(alpha = 0.5, color = "forestgreen") +
  labs(x = "Year",
       y = "Forest Area (sq. km)") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5))
```

### 2. OECD Data (10 pts)


```{r message=FALSE}
print("2a")
oecd <- read_csv("oecd_annual_data.csv", skip = 4, col_names = TRUE)
oecd
```

```{r}
print("2b: last 6 rows")
oecd <- oecd %>% 
  fill(OECD_member)
tail(oecd)

print("2c")
oecd <- oecd %>% 
  pivot_longer(`2000`:`2022`, names_to = "Year", values_to = "MarineProtectedArea_sqkm")
oecd

print("2d")
oecd |> 
  filter(OECD_member == "Non-OECD Economies") |> 
  pivot_wider(names_from = Country, values_from = MarineProtectedArea_sqkm) |> 
  select(-OECD_member)
```

### 3. Santa Cruz Rodents Data Cleaning (20 pts)


```{r message=FALSE}
rodents <- read_csv("capture_data.csv")

#a
rodents <- rodents %>% 
  rename(TrapID = `Trap ID`, Status = `Status (R/N)`, TotalWeight = `Total Weight`,
         BagWeight = `Bag weight`, AnimalWeight = `Animal Weight`, 
         HindfoodLength = `Hind foot length`, TailLength = `Tail length`,
         HairSample = `Hair sample (Y/N)`, Position = `Position (R/L)`)

#b
print("3b")
rodents <- rodents %>% 
  fill(Site)
rodents

#c
rodents <- rodents %>% 
  mutate(Species = replace(Species, Species == "SIOC?", "SIOC"),
         Species = replace(Species, Species == "DIME?", "DIME"))

#d
rodents <- rodents %>% 
  mutate(TailLength = replace(TailLength, TailLength == "~15.5", "15.5"))

#e
rodents <- rodents %>% 
  mutate(HairSample = na_if(HairSample, "?"),
         Position = na_if(Position, "?"))

#f
print("3f")
ggplot(rodents, aes(TotalWeight, fill = Site)) +
  geom_histogram(alpha = 0.5, position = "identity", binwidth = 10) +
  facet_wrap(~ Species) +
  labs(x = "Weight (g)", y = "Frequency") +
  theme_classic()

```


### 4. Remembering Joins (15 pts)

```{r message=FALSE}
veg <- read_csv("microsite_grouped_veg.csv")

#a
veg <- veg %>% 
  rename(RecordID = `...1`, TrapID = `Trap Location`, DistancetoVeg_m = `Distance to Vegetation (m)`,
         VegetationType = `Type of Vegetation`, PercentCover = `Percent Veg Cover`, DistancetoWater_m = `Distance to Water (m)`)

#b
print("4b")
veg <- veg %>% 
  select(Site, TrapID, Grouped_Veg)
veg

#c
print("4c")
inner_join(rodents, veg)
```

### 5. Santa Cruz Rodents Wrangling (20 pts)

```{r}
#a
print("5a")
rodents <- rodents %>% 
  separate(Date, into = c("Year", "Month", "Day"), sep = "-")
rodents

#b
print("5b")
rodents <- rodents %>% 
  unite("Date", Year:Day, sep = "-")
rodents

#c
print("5c")
sp_by_site <- rodents %>% 
  group_by(Site, Species) %>% 
  summarize(Count = n())
sp_by_site

#d
print("5d")
sp_by_site %>% 
  pivot_wider(names_from = Species,
              values_from = Count,
              values_fill = 0)
```

### 6. Mammals (20 pts)

The code chunk below has some made-up mammal data. Run the code chunk below to complete question 5.

```{r}
mammals <- data.frame(site = c(1,1,2,3,3,3), 
                      taxon = c('Suncus etruscus', 'Sorex cinereus', 
                                'Myotis nigricans', 'Notiosorex crawfordi', 
                                'Suncus etruscus', 'Myotis nigricans'),
                      density = c(6.2, 5.2, 11.0, 1.2, 9.4, 9.6),
                      avg_mass = c(4.2, 5, 9.1, 8.6, 4.1, 8.7))
```

```{r}
print("6a")
mammals <- mammals %>% 
  separate(taxon, c("genus", "species"), sep = " ")
mammals

print("6b")
mammals <- mammals %>% 
  pivot_longer(density:avg_mass, names_to = "measurement", values_to = "value")
mammals

print("6c")
mammals <- mammals %>% 
  unite("taxon", genus, species, sep = " ")
mammals

print("6d")
mammals <- mammals %>% 
  pivot_wider(names_from = measurement, values_from = value)
mammals
```
